---
title: "Botanical Value MapWorkflow - Habitat quality analysis"
author: "Natural England"
creation date: "16/03/2022"
last edited: "18/03/2023"
output:
  html_document:
    includes:
      before_body: './images/header.html'
always_allow_html: yes
---

```{r setup, include=F}

knitr::opts_chunk$set(echo = TRUE, eval = F)

#load packages
require(tidyverse)
require(sf)
require(ggpubr)
library(viridis)
library(tmap)
library(furrr)
library(RColorBrewer)

source('HeatMapWorkflow_functions.R')
```

## Project Background

The botanical value map is a categorised map identifying areas of high, moderate and low value for plant species. These are derived from the botanical heatmaps and assess the presence of priority species and positive habitat indicators that confirm the present of good quality semi-natural habitat.This was developed for the purpose of helping to inform screening of tree planting activities and other land management decisions. 

The botanical value of a monad will be determined by two steps:

1. A first screen of sites where priority species have been recorded. These are species which are GB rare, scarce or threatened species. These are given a "High" value status, where any listed conservation species have been recorded within a monad.

2. Habitat quality is then assessed through observing the number of positive habitat indicators present per monad. These indicators have been determined from the BSBI axiophyte list, common standards monitoring and ancient woodland indicator species. The total for the monad has then been compared to the total species indicators present in the surrounding neighbourhood in order to alleviate some of the spatial bias in the data. Different methods for establishing the neighbourhood as well as for the proportional thresholds to use for low, moderate and high categories were trialled.

## Comparison of Habitat quality assessments

The presence of a high quality habitat is assessed with the BSBI heat maps through benchmarking the number of positive habitat indicator species found present within each monad. As indicator numbers vary regionally across England, we looked to find the best way to benchmark high, medium and low categories accounting for this variation. For this we looked at three different methods where benchmarks would be set according to:

* a proportion of the total indicators present within a region
* a proportion of the total indicators present within a myriad (100 x 100 km grid cells)
* a proportion of the total indicators present within a moving neighbourhood of surrounding monads

We also looked at parameters within these and different proportional thresholds to categorise the three value classes.

See technical report (Trippier et al. 2022) for write up of the analysis results.

```{r readHeatmap, eval=F}
#list layers in 1km botanical heatmap geodatabase
st_layers(paste0(outputFolder,'/BotanicalHeatMaps_Indicators_1970_2021.gpkg'))

# read in heatmap monad data
mainTable <- st_read(dsn=paste0(outputFolder,'/BotanicalHeatMaps_Indicators_1970_2021.gpkg'),layer='BotanicalIndicators_England_1km')
all_species <- st_read(dsn=paste0(outputFolder,'/BotanicalHeatMaps_Indicators_1970_2021.gpkg'),layer='BotanicalIndicators_England_species_1km')
species_hab <- st_read(dsn=paste0(outputFolder,'/BotanicalHeatMaps_Indicators_1970_2021.gpkg'),layer='broadHabitat_associated_species')

```

### 1. Analysis of regional indicator benchmarking

```{r regionalProps, eval=F}
#load in data layers
all_monads <- mainTable
# find all habitats
habs <- names(all_monads)[8:17]
#match column headers
names(species_hab)[4:13]<- names(all_monads)[8:17]

#find all regions
region_cat <- unique(all_monads$region)

#iterate through habitats for regional assessment
all_hab_summary <- NULL
for (i in habs){
  print(i)
  hab_monad <- all_monads %>% select(monad, region, total_no=!!i) %>%
    mutate(total_no = ifelse(total_no==-9999, NA, total_no))
  #join all species with species hab lookup
  hab_species <- all_species %>% left_join(species_hab[,c('species',i)], by='species')  %>% filter(get(i) == 1) %>%  select(monad, species, lastRecorded)
  hab_all <- inner_join(hab_monad,hab_species,by='monad') 
  #unique species
  total_species <- hab_all %>% st_drop_geometry() %>% select(region,total_no,species)
  #iterate through regions to calculate total, 10% and 20%
  all_regions <- purrr::map_df(region_cat,.f=function(cat){
    hab_region <- hab_monad %>% filter(region==cat) 
    #total number indicators present in a region
    Tot_indicators <- total_species %>% filter(region==cat) %>% 
    dplyr::select(species) %>% unique() %>% nrow()
    #10% indicators present = moderate, 20% =good
    pcnt_5 <- round(0.05*Tot_indicators)
    pcnt_10 <- round(0.1*Tot_indicators)
    pcnt_20 <- round(0.2*Tot_indicators)
    quantile(Tot_indicators)
    hab_region <- hab_region %>% mutate(Tot_indicators=Tot_indicators,
                                        pcnt5 = round(0.05*Tot_indicators), 
                                        pcnt10= round(0.1*Tot_indicators),
                                        pcnt15= round(0.15*Tot_indicators),
                                        pcnt20= round(0.2*Tot_indicators),
                                        pcnt25= round(0.25*Tot_indicators),
                                        pcnt30= round(0.3*Tot_indicators),
                                        pcnt35= round(0.35*Tot_indicators),
                                        pcnt40= round(0.4*Tot_indicators),
                                        pcnt45= round(0.45*Tot_indicators),
                                        pcnt50= round(0.5*Tot_indicators)
                                       )
    hab_region
  })
  # regional differences in thresholds
  #histogram
  p <- ggplot(data = all_regions, aes(x = total_no)) + geom_histogram() 
  p + facet_wrap(~region, scales = "free_y") + xlab("Number of indicator species") + ylab("Recorder frequency")
  ggsave(paste0('IndicatorBaselines/Regional/',i,'_regionalhistograms.png')) 
  
  #regional differences in percentiles
  region_summ <- all_regions %>% st_drop_geometry() %>% 
    select(region, Tot_indicators:pcnt50) %>% unique() 
  region_graph <- region_summ %>% gather(key = "proportion", value = "value",Tot_indicators:pcnt50)
  region_graph$proportion <- factor(region_graph$proportion,
                        levels=c("Tot_indicators","pcnt50","pcnt45","pcnt40","pcnt35","pcnt30","pcnt25","pcnt20","pcnt15","pcnt10","pcnt5",ordered=TRUE))
  
  
  #proportions plot
  region_graph <- region_graph %>% filter(proportion=="pcnt5"|proportion=="pcnt10"|proportion=="pcnt20"|proportion=="pcnt30"|proportion=="Tot_indicators")
  propPlot <- ggplot(region_graph, aes(x = region, y = value, fill = proportion)) +
    geom_bar(position="dodge", stat="identity", show.legend = T) +
    scale_x_discrete(guide = guide_axis(angle = 90),labels = function(x) str_wrap(x, width = 10)) +
    scale_discrete_manual(aesthetics = 'fill', values = viridis(12), name="Proportion of indicators",labels=c("100%","30%", "20%","10%","5%")) + 
    xlab("Region") + ylab("Number of indicator species") + ggtitle(i) +
    theme(legend.position="top")
  #summary table
  summ_tab<- region_summ %>% dplyr::select(region,p0.05=pcnt5,p0.1=pcnt10,p0.15=pcnt15, p0.2=pcnt20,p0.25=pcnt25,p0.3=pcnt30,p0.35=pcnt35,p0.4=pcnt40,p0.45=pcnt45,p0.5=pcnt50,p1=Tot_indicators) 
  ggsumtab <- summ_tab %>% ggtexttable(rows = NULL,theme = ttheme("mOrange",base_size = 9))
  #grid table and graph
  multi <- ggarrange(propPlot, ggsumtab,
                     ncol = 1, nrow = 2,
                     widths= c(0.1,0.8,0.1),
                     heights = c(2, 1)) +
    theme(plot.margin = margin(1,1,1,1, "cm"))
  ggsave <- ggplot2::ggsave; body(ggsave) <- body(ggplot2::ggsave)[-2]
  ggsave(paste0('IndicatorBaselines/Regional/Proportions/',i,'_summary.pdf'))
  #add to end table
  summ_tab<- summ_tab %>% mutate(habitat=i)
  all_hab_summary <- rbind(all_hab_summary,summ_tab)
}

write.csv(all_hab_summary,'IndicatorBaselines/Regional/all_hab_summary.csv')
```

```{r regionalClassifying, eval=F}
#classifying habitats into traffic light categories
all_hab_summary <-read.csv('IndicatorBaselines/Regional/all_hab_summary.csv')

# iterate through testing different regional percentage benchmarks for the categorised layers
tests <- tribble(  ~test, ~cat,~max,
                   1,'low', 0.1,
                   1,'moderate', 0.25,
                   1,'high', 1,
                   2,'low', 0.1,
                   2,'moderate', 0.2,
                   2,'high', 1,
                   3,'low', 0.05,
                   3,'moderate', 0.1,
                   3,'high', 1,
                   4,'low', 0.1,
                   4,'moderate', 0.3,
                   4,'high', 1,
                   )
for (i in unique(tests$test)){
  print(i)
  test_select <- tests %>% filter(test==i)
  outfold_test <- paste0('IndicatorBaselines/Regional/CategorisedVectors/BotanicalVal_',i,'_perc_',test_select$max[1], '_',test_select$max[2])
  dir.create(outfold_test)
  for (j in habs){
  classifyHabs(habType = j, levels = test_select[,c('cat','max')], outfolder=paste0(outfold_test,'/'))
  }
}

```

```{r regionMapAnalysis,  eval = F}
#extract results for each test
tests <- basename(list.dirs('IndicatorBaselines/Regional/CategorisedVectors/', recursive = F))

for(i in 1:length(tests)){
  testname <- tests[i]
  files <- list.files(paste0('IndicatorBaselines/Regional/CategorisedVectors/',testname,'/'),  pattern='shp',full.names = T)
  all_out <- purrr::map_df(files,.f=function(file){
    map_out <-st_read(file,quiet=T)
    #count monads in each category
    map_out %>% st_drop_geometry() %>% group_by(valueCat) %>% summarise(count = n())
    #by region
    result <- map_out %>% st_drop_geometry()  %>% 
      group_by(valueCat,region) %>% summarise(count = n()) %>% 
      mutate(habitat=str_remove(basename(file),'_regional.shp'))
    result
  })
  #as factor
  all_out$valueCat[is.na(all_out$valueCat)]<-'Further Survey Required'
  all_out$valueCat <- factor(all_out$valueCat,
                             levels=c("high","moderate","low",'Further Survey Required',ordered=TRUE))
  #summary across all regions
  by_hab <- all_out %>% group_by(habitat,valueCat) %>% summarise(count = sum(count)) %>%
    pivot_wider(names_from=valueCat,values_from=count)
  
  #plot in 2 pages
  all_out1<- all_out %>% filter(habitat %in% unique(all_out$habitat)[1:6])
  all_out2<- all_out %>% filter(habitat %in% unique(all_out$habitat)[7:10])
  allhabs <- list(all_out1,all_out2)
  pal <- c('#fde725','#5ec962','#21918c','#DCDCDC')
  #proportions plot
  for(i in 1:2){
    habset <- allhabs[[i]]
    if(i==1){
      labs <- c('arable','boundary','coastal','fen, marsh, swamp','freshwater','grassland')
      names(labs) <- c('arable','boundry','coastal','fnMrshS','frshwtr','grsslnd')
    }else{
      labs <- c('heath & bog','inland rock','montane','woodland')
      names(labs) <- c('heathBg','inlndRc','montane','woodlnd')
    }
    ggplot(habset, aes(x = region, y = count, fill = valueCat)) +
      geom_bar(position="fill", stat="identity", show.legend = T) +
      scale_x_discrete(guide = guide_axis(angle = 90),labels = function(x) str_wrap(x, width = 10)) +
      scale_discrete_manual(aesthetics = 'fill', values = pal, name="Botanical value",labels=c("High","Moderate", "Low",'Further survey required')) + 
      xlab("Region") + ylab("Proportion of regional monads") + theme_bw() + theme(legend.position="bottom") +  
      facet_wrap(~ habitat, ncol=2, labeller = labeller(habitat = labs))
    ggsave(paste0('IndicatorBaselines/Regional/CategorisedVectors/',testname,'_habset',i,'.png'))
  }
}

```

### 2. Analysis of myriad indicator benchmarking

```{r myriadanalysis, eval = F}
#read in OS grids
Ind_monads <- mainTable
myriads <- st_read('DataLayers/OS BNG grids/OS_BNG_grids_100km_grid.shp') %>% select(tile_name)

#join to myriads
monadmyriads <- Ind_monads  %>% st_centroid() %>% st_intersection(myriads) %>%  st_drop_geometry() %>% select(monad, myriads=tile_name)
Ind_monad_myriad <- Ind_monads %>% left_join(monadmyriads, by="monad")

# find all habitats
habs <- names(Ind_monad_myriad)[8:17]
#match column headers
names(species_hab)[4:13]<- names(Ind_monad_myriad)[8:17]

#iterate through trials of different thresholds
tests <- tribble(  ~test, ~cat,~max,
                   1,'low', 0.1,
                   1,'moderate', 0.2,
                   1,'high', 1,
                   2,'low', 0.05,
                   2,'moderate', 0.1,
                   2,'high', 1
                   )
for (i in unique(tests$test)){
  print(i)
  test_select <- tests %>% filter(test==i)
  outfold_test <- paste0('IndicatorBaselines/Tetrad/BotanicalVal_',i,'_perc_',test_select$max[1], '_',test_select$max[2])
  dir.create(outfold_test)
  for (j in habs){
  classifyTetHabs(shapefile= Ind_monad_myriad, habType = j, levels = test_select[,c('cat','max')], outfolder=paste0(outfold_test,'/'))
  }
}


```

```{r myriadMapAnalysis, eval = F }
#extract results
testname <- 'BotanicalVal_1_perc_0.1_0.2'

files <- list.files(paste0('IndicatorBaselines/MovingWindow/',testname,'/'),  pattern='shp',full.names = T)
map <- st_read(files) %>% st_drop_geometry()
map  %>% group_by(valueCat) %>% summarise(count = n())

```

### Analysis of  indicator thresholds within a moving window of neighbouring hectads

* note this requires a lot of processing power! may take a while to run

```{r movingWindow, eval = F}
#read in OS grids
Ind_monads <- mainTable
#match column headers
names(species_hab)[4:13]<- names(Ind_monads)[8:17]

#distance function - 10000 = 1 hectad, 5000 = centroid to edge
neighbour_dist<-function(x,unit=1000){x*unit - (unit/2)}

# load habitat names
habs <- names(Ind_monads)[8:17]
habType ='frshwtr' # or set to singular

# trial different regional percentage benchmarks for the categorised layers - distances 50 and 25
tests <- tribble(  ~test, ~cat,~max,
                   1,'low', 0.1,
                   1,'moderate', 0.2,
                   1,'high', 1,
                   2,'low', 0.1,
                   2,'moderate', 0.2,
                   2,'high', 1,
                   )
for (i in unique(tests$test)){
  print(i)
  test_select <- tests %>% filter(test==i)
  outfold_test<- paste0('IndicatorBaselines/MovingWindow/BotanicalVal_',i,'_perc_',test_select$max[1], '_',test_select$max[2])
  dir.create(outfold_test)
  for (j in habs[1:2]){
  classifyMovingWindow(habType = j, levels = test_select[,c('cat','max')], outfolder=paste0(outfold_test,'/'))
    print(paste(j, ' done.')) 
  }
}

```

```{r movingwindowAnalysis, eval = F}
#extract results
testname <- 'MovingWindow_50km_perc_0.1_0.2'

files <- list.files(paste0('IndicatorBaselines/Tetrad/',testname,'/'),  pattern='shp',full.names = T)

all_out <- purrr::map_df(files,.f=function(file){
  map_out <-st_read(file,quiet=T)
  #count monads in each category
  map_out %>% st_drop_geometry() %>% group_by(valueCat) %>% summarise(count = n())
  #by region
  result <- map_out %>% st_drop_geometry()  %>% 
    group_by(valueCat,tetrad) %>% summarise(count = n()) %>% 
    mutate(habitat=str_remove(basename(file),'_regional.shp'))
result
})

#as factor
all_out$valueCat[is.na(all_out$valueCat)]<-'Further Survey Required'
all_out$valueCat <- factor(all_out$valueCat,
                        levels=c("high","moderate","low",'Further Survey Required',ordered=TRUE))
#summary across all regions
by_hab <- all_out %>% group_by(habitat,valueCat) %>% summarise(count = sum(count)) %>%
  pivot_wider(names_from=valueCat,values_from=count)

  #proportions plot
  labs <- c('arable','boundary','coastal','fen, marsh, swamp','freshwater','grassland','heath & bog','inland rock','montane','woodland')
  names(labs) <- c('arable','boundry','coastal','fnMrshS','frshwtr','grsslnd','heathBg','inlndRc','montane','woodlnd')
pal=rev(brewer.pal(4,'Greys')) 
 ggplot(all_out, aes(x = habitat, y = count, fill = valueCat)) +
    geom_bar(position="fill", stat="identity", show.legend = T) +
    scale_x_discrete(labels = str_wrap(labs, width = 10)) +
    scale_discrete_manual(aesthetics = 'fill', values = pal, name="Botanical value",labels=c("High","Moderate", "Low",'Further survey required')) + 
    xlab("") + ylab("Proportion of monads") + theme_bw() + theme(legend.position="bottom")
 ggsave(paste0('IndicatorBaselines/Tetrad/',testname,'.png'), width = 8, height=4)

```

## Combining the steps into a the botanical value layer

The resulting values from the first and second steps were then combined to take the greatest value level assigned to a monad, therefore representing a precautionary approach to preserve those sites which are potentially of high botanical value.

```{r botanicalValue, eval = F }

#locations for regional method layers
methods <- list(regions='IndicatorBaselines/Regional/CategorisedVectors/BotanicalVal_2_perc_0.1_0.2_v2/',
                tetrad='IndicatorBaselines/Tetrad/BotanicalVal_1_perc_0.1_0.2/')

#iterate producing final layers for the different methods
for (i in methods){
  files <-list.files(unlist(i), pattern='.shp')
  priority <- mainTable %>% select(monad, ttPrrty) %>% mutate(priorityCat = ifelse(ttPrrty==-9999,'NA',
                                                                              ifelse(ttPrrty>=1,'high','NA'))) %>% select(-ttPrrty)
for (j in files){
  layer <- st_read(paste0(unlist(i),j),quiet=T)
  hab <- str_replace(j,'_regional.shp','')
  layer <- layer %>% st_drop_geometry() %>% select(monad,{{hab}}:=valueCat)
  priority <- priority %>% left_join(layer, by='monad')
  print(paste(str_replace(j,'_regional.shp',''), 'done.'))
}

  #start with priority and update with highest rank - for loop gets stuck here but runs manually
  valueTable <- priority %>% replace(is.na(.),'further survey required')
valueTable <- valueTable %>% 
  dplyr::mutate(valueCat = ifelse(priorityCat=='high'|arable=='high' |boundry =='high' |coastal =='high' |fnMrshS=='high' |frshwtr=='high'|grsslnd=='high' | heathBg=='high' |inlndRc =='high' |montane =='high' |woodlnd =='high','high',
                           ifelse(arable=='moderate' |boundry =='moderate' |coastal =='moderate' |fnMrshS=='moderate' |frshwtr=='moderate'|grsslnd=='moderate' | heathBg=='moderate' |inlndRc =='moderate' |montane =='moderate' |woodlnd =='moderate','moderate',
                                  ifelse(arable=='low' |boundry =='low' |coastal =='low' |fnMrshS=='low' |frshwtr=='low'|grsslnd=='low' | heathBg=='low' |inlndRc =='low' |montane =='low' |woodlnd =='low','low', 'further survey required'))))

#save layer
st_write(valueTable,paste0('IndicatorBaselines/botanicalValue_',names(i),'_England_1km.shp'), delete_layer = T)


##plot ##
plotTable <- valueTable %>% mutate(valueCat =  factor(valueCat,levels=c('further survey required',"low","moderate","high"))) 
pal <- c('#DCDCDC','#bae4b3','#74c476','#238b45')

map1 <- tm_shape(plotTable) + tm_fill(col='valueCat', palette = pal, reverse=T,title = str_wrap(paste0('Botanical value'),width=10)) + tm_scale_bar(position=c("left", "bottom"))
tmap_save(map1, paste0('IndicatorBaselines/botanicalValue_',names(i),'_England_1km.png'))  

#just habitat quality metrics
habValueTable <- valueTable %>% 
  dplyr::mutate(habQual = ifelse(arable=='high' |boundry =='high' |coastal =='high' |fnMrshS=='high' |frshwtr=='high'|grsslnd=='high' | heathBg=='high' |inlndRc =='high' |montane =='high' |woodlnd =='high','high',
                                  ifelse(arable=='moderate' |boundry =='moderate' |coastal =='moderate' |fnMrshS=='moderate' |frshwtr=='moderate'|grsslnd=='moderate' | heathBg=='moderate' |inlndRc =='moderate' |montane =='moderate' |woodlnd =='moderate','moderate',
                                         ifelse(arable=='low' |boundry =='low' |coastal =='low' |fnMrshS=='low' |frshwtr=='low'|grsslnd=='low' | heathBg=='low' |inlndRc =='low' |montane =='low' |woodlnd =='low','low', 'further survey required'))))

##plot ##                                                  
plotTable <- habValueTable %>% mutate(habQual =  factor(habQual,levels=c('further survey required',"low","moderate","high"))) 
pal <- c('#DCDCDC','#bae4b3','#74c476','#238b45')

map2 <- tm_shape(plotTable) + tm_fill(col='habQual', palette = pal, reverse=T,title = str_wrap(paste0('Botanical value'),width=10)) + tm_scale_bar(position=c("left", "bottom"))
tmap_save(map2, paste0('IndicatorBaselines/habitatQuality_',names(i),'_England_1km.png'))  
}

```

Analysis of output botanical value map classified monads.

```{r BVAnalysis, eval = F }
#extract results from each analysis
files <- list.files(paste0('IndicatorBaselines/'),  pattern='shp',full.names = T)

all_out <- purrr::map_df(files,.f=function(file){
  map_out <-st_read(file,quiet=T)
  
  #count monads in each category
  results <- map_out %>% 
    st_drop_geometry() %>% 
    group_by(valueCt) %>% 
    summarise(count = n())
  
  total <- sum(results$count)
  results <- results %>% 
    mutate(prop = round((count/total)*100,2)) %>%
    mutate(test = unlist(str_split(basename(file),'_'))[2])
  results

})

##just moving window calculate area
mw_map <- st_read(files[1])
england <- st_read('DataLayers/Countries_(December_2020)_UK_BFC/England_boundary.shp')
#clip to england coastline
mw_coast <- st_intersection(mw_map, england)
st_write('IndicatorBaselines/MovingWindow/mw_England_botanicalValue_clipped.shp')

#merge
library(units)

mw_coast <-st_read('IndicatorBaselines/MovingWindow/mw_England_botanicalValue_clipped.shp')

mw_area <- mw_coast %>% select(monad,valueCt) %>%
  mutate(area_m=st_area(geometry)) %>%
  st_drop_geometry() %>%
  mutate(area_km= set_units(area_m, km^2)) %>%
  group_by(valueCt) %>% 
  summarise(totArea_km = sum(area_km))


#sense check - area england 130,279 km²
sum(mw_area$totArea_km)
```
