---
title: "Botanical Value Map Workflow"
author: "Natural England"
date: "28/04/2021"
output:
  html_document:
    includes:
      before_body: 'D:/Github/BotanicalHeatmaps/images/header.html'
always_allow_html: yes
---

```{r setup, include=F}

knitr::opts_chunk$set(echo = TRUE, eval=F)

#load packages
require(tidyverse)
require(sf)
require(ggpubr)
library(viridis)
library(tmap)
library(furrr)
library(RColorBrewer)
library(knitr)
#devtools::install_github("jabiru/tictoc")
library(tictoc)

source('HeatMapWorkflow_functions.R')
```

## Project Background

The botanical value map is a categorised map identifying areas of high, moderate and low value for plant species. These are derived from the botanical heatmaps and assess the presence of Rare, Scarce or Threatened (RST) plant species and positive habitat priority indicators (PHPI) species that confirm the present of good quality semi-natural habitat. This was developed for the purpose of helping to inform screening of tree planting activities and other land management decisions. 

The botanical value of a monad will be determined by two steps:

1. A first screen of sites where RST plant species have been recorded. These are species which are GB rare, scarce or threatened species. These are given a "High" value status, where any listed priority species have been recorded within a monad between 1970 and 2021.

2. Habitat quality is then assessed through observing the number of PHPI species present per monad. These indicators have been determined from the BSBI axiophyte list, common standards monitoring and ancient woodland indicator species. The total for the monad has then been compared to the total species indicators present in the surrounding neighborhood in order to alleviate some of the spatial bias in the data. Different methods for establishing the neighborhood as well as for the proportional thresholds to use for low, moderate and high categories were trialed.


```{r filepathsToAdd}
# Folder path to save outputs
outputFolder <- 'Y:/BotanicalHeatmaps/Analysis/2021-22/Products Final2/Data/'

# Folder path to save outputs
outputFolder <- 'Y:/BotanicalHeatmaps/Analysis/2022-23/Outputs/'
```

```{r readHeatmap}
#list layers in 1km botanical heatmap geodatabase
st_layers(paste0(outputFolder,'/BotanicalHeatMaps_Indicators_1970_2022.gpkg'))

# read in heatmap monad data
mainTable <- st_read(dsn=paste0(outputFolder,'/BotanicalHeatMaps_Indicators_1970_2022.gpkg'),layer='BotanicalIndicators_England_1km')
all_species <- st_read(dsn=paste0(outputFolder,'/BotanicalHeatMaps_Indicators_1970_2022.gpkg'),layer='PHPI_England_species_1km')
species_hab <- st_read(dsn=paste0(outputFolder,'/BotanicalHeatMaps_Indicators_1970_2022.gpkg'),layer='broadHabitat_associated_species')

```

## Step 1: Presence of priority species

High = 1 or more priority species present within a monad
Low = 0 priority species present within a monad of good survey coverage
Further Survey Required = 0 priority species present within a monad of poor survey coverage

```{r priorityCheck, eval=T}
# select data and categorise
priority <- mainTable %>% select(monad, RSTsp,region) %>% mutate(valueCat = ifelse(RSTsp==-9999,'No indicators, poor coverage',ifelse(RSTsp>=1,'High','No indicators, good coverage')))
#count monads in each category
countdf <- priority %>% st_drop_geometry() %>% group_by(valueCat) %>% summarise(count = n())
```

* Percentage cover of High monads: `r (countdf$count[countdf$valueCat=='High']/nrow(mainTable))*100`
* Pecentage cover Low monads: `r (countdf$count[countdf$valueCat=='Low']/nrow(mainTable))*100`
* Pecentage cover Further Survey Required monads: `r (countdf$count[countdf$valueCat=='Further Survey Required']/nrow(mainTable))*100`

```{r prioritytable, eval=T}
#table of values by region
priortytab <- priority %>% st_drop_geometry() %>% group_by(valueCat,region) %>% summarise(count = n())
knitr::kable(priortytab)
```

```{r priorityCheckPlot, eval=T}
#convert to a factor
hab_monad <- priority %>% mutate(valueCat =  factor(valueCat,levels=c("No indicators, poor coverage","No indicators, good coverage","High"))) 

# create map plot
pal <- c('#afafaf','#f8f8fa', "#21908CFF")
map1 <- tm_shape(hab_monad) + 
  tm_fill(col='valueCat', style='cat',palette = pal, reverse=T, title=str_wrap('Number of RST plant species recorded',width=40)) + 
  tm_scale_bar(position=c("right", "bottom")) #+  
 # tm_credits(str_wrap('Data were provided by BSBI. Contains OS data © Crown copyright and database right 2021. Source: Office for National Statistics licensed under the Open Government Licence v.3.0.', width=90), position= 'left', size=0.6) 

tmap_save(map1, paste0(outputFolder,'/priorityIndicatorsMap.png'))  

```

## Step 2: Presence of a high quality habitat

The presence of a high quality habitat is assessed with the BSBI heat maps through benchmarking the number of PHPI species found present within each monad. As indicator numbers vary regionally across England, we looked to find the best way to benchmark high, medium and low categories accounting for this variation. For this we looked at three different methods where benchmarks would be set according to:

* a proportion of the total indicators present within a region
* a proportion of the total indicators present within a myriad
* a proportion of the total indicators present within a moving neighbourhood of surrounding monads

See 'habitat_quality_analysis.Rmd' for further information on this analysis. 

After consultation with the habitat specialists we concluded that the neighbouring window analysis would be the best approach using a distance radius of 25km.

### Analysis of  indicator thresholds within a moving window of neighbouring hectads

* note this requires a lot of processing power! may take a while to run

```{r movingWindow, eval = F}
#read in OS grids
Ind_monads <- mainTable

#match column headers - if updating check these line up and follow same naming convention
names(species_hab)[3:12]<- names(Ind_monads)[9:18]

#distance function - 1000 = 1 monad
neighbour_dist<-function(x,unit=1000){x*unit - (unit/2)}

### run for one habitat group ###
# load habitat names
habs <- names(Ind_monads)[9:18]
habType ='grassland'

## function - classify by moving monad window
#classifyMovingWindow(Ind_monads,region = 'monad',habType,levels = data.frame(cat=c("low","moderate","high"),max= c(0.1,0.2,1)),
#                         movingDist = neighbour_dist(25),outfolder =paste0(outputFolder,'/HabitatQualityMW/' ))

### loop through habitat groups ###
#list habitat classes
habList <- habs[1:7]
#iterate through
for (i in 1:length(habList)){
  tic(paste(i,"processing"))
  print(paste(habList[i], 'started..'))
  habType = habList[i]
  classifyMovingWindow(Ind_monads,region = 'monad',
                       habType,
                       levels = data.frame(cat=c("low","moderate","high"),max= c(0.1,0.2,1)),
                       movingDist = neighbour_dist(25),
                       outfolder =paste0(outputFolder,'/HabitatQualityMW/' ))
  print(paste(habList[i], 'done.'))
  toc()
}

```

```{r habitatQualityAnalysis,eval=F }
#extract resulting files
files <- list.files(paste0(outputFolder,'/HabitatQualityMW/'),  pattern='shp',full.names = T)

#iterate reading in the files created by moving window analysis
all_out <- purrr::map_df(files,.f=function(file){
  map_out <-st_read(file,quiet=T)
  #count monads in each category
  map_out %>% st_drop_geometry() %>% group_by(valueCat) %>% summarise(count = n())
  #by region
  result <- map_out %>% st_drop_geometry()  %>% 
    group_by(valueCat) %>% summarise(count = n()) %>% 
    mutate(habitat=str_remove(basename(file),'_mw.shp'))
result
})

#check any NAs
all_out$valueCat[is.na(all_out$valueCat)]
#as factor
all_out$valueCat <- factor(all_out$valueCat,
                        levels=c("High","Moderate","Low",'no indicators,good survey coverage','no indicators, poor survey coverage',ordered=TRUE))

#summary across all habitats
by_hab <- all_out %>% group_by(habitat,valueCat) %>% summarise(count = sum(count)) %>%
  pivot_wider(names_from=valueCat,values_from=count)

  #proportions plot
  labs <- c('arable','boundary','coastal','fen, marsh, swamp','freshwater','grassland','heath & bog','inland rock','montane','woodland')
  names(labs) <- c('arable','boundary','coastal','fenMarshSwamp','freshwater','grassland','heathBog','inlandRock','montane','woodland')
pal=rev(brewer.pal(5,'Greys')) 
 ggplot(all_out, aes(x = habitat, y = count, fill = valueCat)) +
    geom_bar(position="fill", stat="identity", show.legend = T) +
    scale_x_discrete(labels = str_wrap(labs, width = 10)) +
    scale_discrete_manual(aesthetics = 'fill', values = pal, name="Botanical value",labels=c("High","Moderate", "Low",'no indicators,good survey coverage','no indicators, poor survey coverage')) + 
    xlab("") + ylab("Proportion of monads") + theme_bw() + theme(legend.position="bottom")
 ggsave(paste0(outputFolder,'HabitatQualityMW/HabitatQualityMW25km.png'), width = 8, height=4)

```


## Combining the steps into a the botanical value layer

The resulting values from the first and second steps were then combined to take the greatest value level assigned to a monad, therefore representing a precautionary approach to preserve those sites which are potentially of high botanical value.

```{r botanicalValue}

#get habitat quality files
files <- list.files(paste0(outputFolder,'/HabitatQualityMW/'),  pattern='shp',full.names = T)

#get priority category values
priority <- mainTable %>% select(monad,surveyCov, RSTsp) %>% 
  mutate(RSTCat = ifelse(RSTsp==-9999,'no indicators, poor survey coverage',
                              ifelse(RSTsp>=1,'High','no indicators, good survey coverage'))) %>% 
  select(-RSTsp)
unique(priority$RSTCat)

#get habitat counts
habCounts <- mainTable %>% select(monad,araCount=arable,bouCount=boundary,couCount=coastal,fenCount=fenMarshSwamp,freCount=freshwater,graCount=grassland,heaCount=heathBog,inlCount=inlandRock,monCount=montane,wooCount=woodland)

#get broad habitat category values
for (j in files){
  layer <- st_read(j,quiet=T)
  hab <- str_replace(basename(j),'_mw.shp','')
  layer <- layer %>% st_drop_geometry() %>% select(monad,{{hab}}:=valueCat)
  habCounts <- habCounts %>% left_join(layer, by='monad')
  print(paste(str_replace(basename(j),'_mw.shp',''), 'done.'))
}
#sense check values and categories

habJoin <- habCounts %>% select(monad,arable,boundary,coastal,fenMarshSwamp,freshwater,grassland,heathBog,inlandRock,montane,woodland) %>% st_drop_geometry()

#compile table
valueTable <- priority %>% left_join(habJoin,by='monad') 
names(valueTable) <- c('monad','surveyCov','RSTsp','arable','boundary','coastal','fenMarshSwamp','freshwater','grassland','heathBog','inlandRock','montane','woodland','geom')

#start with priority and update with highest rank 
valueTable <- valueTable %>% 
  dplyr::mutate(valueCat = ifelse(RSTsp=='High'|arable=='High' |boundary =='High' |coastal =='High' |fenMarshSwamp=='High' |freshwater=='High'|grassland=='High' | heathBog=='High' |inlandRock =='High' |montane =='High' |woodland =='High','High',
                           ifelse(arable=='Moderate' |boundary =='Moderate' |coastal =='Moderate' |fenMarshSwamp=='Moderate' |freshwater=='Moderate'|grassland=='Moderate' | heathBog=='Moderate' |inlandRock =='Moderate' |montane =='Moderate' |woodland =='Moderate','Moderate',
                                  ifelse(arable=='Low' |boundary =='Low' |coastal =='Low' |fenMarshSwamp=='Low' |freshwater=='Low'|grassland=='Low' | heathBog=='Low' |inlandRock =='Low' |montane =='Low' |woodland =='Low','Low',
                                         ifelse(arable=='no indicators, good survey coverage' |boundary =='no indicators, good survey coverage' |coastal =='no indicators, good survey coverage' |fenMarshSwamp=='no indicators, good survey coverage' |freshwater=='no indicators, good survey coverage'|grassland=='no indicators, good survey coverage' | heathBog=='no indicators, good survey coverage' |inlandRock =='no indicators, good survey coverage' |montane =='no indicators, good survey coverage' |woodland =='no indicators, good survey coverage','no indicators, good survey coverage',
                                                'no indicators, poor survey coverage')))))
#checks
unique(valueTable$valueCat)
valueTable %>% filter(arable =='no indicators, good survey coverage')

#save layer
st_write(valueTable,paste0(outputFolder,'/botanicalValue_2021_England_1km.shp'), delete_layer = T)


##plot ##
unique(valueTable$valueCat)
plotTable <- valueTable %>%
  mutate(valueCat =  factor(valueCat,levels=c('no indicators, poor survey coverage',"Low","Moderate","High"))) 
pal <- c('#DCDCDC','#bae4b3','#74c476','#238b45')

map1 <- tm_shape(plotTable) + tm_fill(col='valueCat', palette = pal, reverse=T,title = str_wrap(paste0('Botanical value'),width=10)) + tm_scale_bar(position=c("right", "bottom"))
tmap_save(map1, paste0(outputFolder,'/botanicalValue_2021_England_1km.png')) 

```

```{r habqualValue}
# assess for just the habitat quality metrics - priority

habValueTable <- valueTable %>% 
  dplyr::mutate(habQual = ifelse(arable=='High' |boundary =='High' |coastal =='High' |fenMarshSwamp=='High' |freshwater=='High'|grassland=='High' | heathBog=='High' |inlandRock =='High' |montane =='High' |woodland =='High','High',
                                  ifelse(arable=='Moderate' |boundary =='Moderate' |coastal =='Moderate' |fenMarshSwamp=='Moderate' |freshwater=='Moderate'|grassland=='Moderate' | heathBog=='Moderate' |inlandRock =='Moderate' |montane =='Moderate' |woodland =='Moderate','Moderate',
                                         ifelse(arable=='Low' |boundary =='Low' |coastal =='Low' |fenMarshSwamp=='Low' |freshwater=='Low'|grassland=='Low' | heathBog=='Low' |inlandRock =='Low' |montane =='Low' |woodland =='Low','Low',
                                                ifelse(arable=='no indicators, good survey coverage' |boundary =='no indicators, good survey coverage' |coastal =='no indicators, good survey coverage' |fenMarshSwamp=='no indicators, good survey coverage' |freshwater=='no indicators, good survey coverage'|grassland=='no indicators, good survey coverage' | heathBog=='no indicators, good survey coverage' |inlandRock =='no indicators, good survey coverage' |montane =='no indicators, good survey coverage' |woodland =='no indicators, good survey coverage','no indicators, good survey coverage',
                                                'no indicators, poor survey coverage')))))

##plot ##                                                  
plotTable <- habValueTable %>% mutate(habQual =  factor(habQual,levels=c('no indicators, poor survey coverage',"Low","Moderate","High"))) 
pal <- c('#DCDCDC','#bae4b3','#74c476','#238b45')

map2 <- tm_shape(plotTable) + tm_fill(col='habQual', palette = pal, reverse=T,title = str_wrap(paste0('Botanical value'),width=10)) + tm_scale_bar(position=c("left", "bottom"))
tmap_save(map2, paste0(outputFolder,'/habitatQualitymw_2021_England_1km.png'))  


```

Resulting botanical value map stats

```{r BVAnalysis}

#extract results for each of priority species,broad habitat and overall value
allResults <- purrr::map_df(names(valueTable)[c(3:13,15)],.f=function(col){
  #count monads in each category
  results <- valueTable %>% 
    st_drop_geometry() %>%
    select(monad,{{col}}) %>%
    rename(cat = {{col}}) %>% 
    group_by(cat) %>% 
    summarise(count = n())
  #calculate as a percentage of monads
  total <- sum(results$count)
  results <- results %>% 
    mutate(prop = round((count/total)*100,2)) %>%
    mutate(category = col)
  results

})
write.csv(allResults,paste0(outputFolder,'botanicalValueSummary.csv'))
```

```{r BV AreaCalc,eval=F}
##calculate area for final categories
mw_map <- st_read(paste0(outputFolder,'/botanicalValue_2021_England_1km.shp'))
england <- st_read('England_boundary.shp')

#clip monads to england coastline area
mw_coast <- st_intersection(mw_map, england)
mw_coast <-st_write('IndicatorBaselines/MovingWindow/mw_England_botanicalValue_clipped.shp')
```

```{r BV Area,eval=T}
#calculate areas
mw_coast <-st_read('IndicatorBaselines/MovingWindow/mw_England_botanicalValue_clipped.shp')

library(units)
mw_area <- mw_coast %>% select(monad,valueCt) %>%
  mutate(area_m=st_area(geometry)) %>%
  st_drop_geometry() %>%
  mutate(area_km= set_units(area_m, km^2)) %>%
  group_by(valueCt) %>% 
  summarise(totArea_km = sum(area_km))
mw_area
#sense check - area england 130,279 km²
sum(mw_area$totArea_km)
```

## Limitations 

There are several assumptions made throughout the analysis which are key to consider when using these data: 

*	These maps provide an indication of areas where botanically valuable habitats are present to help guide tree planting decisions. However, there are a number of caveats that need to be taken into account when using them. Although detailed, they are not comprehensive, relying on data collected by volunteers often at a coarse capture resolution. Our analysis also makes assumptions on what constitutes ‘good’ recording coverage for a monad. When using these maps for the purpose of tree screening, they should also be consulted alongside other environmental datasets to assess site suitability. 
*	This analysis uses recorder days as a proxy for survey coverage. This is a measure of the number of days where more than 40 species have been recorded during a single visit. However, there is regional bias in species diversity decreasing from south to north. Similarly, many coastal monads contain little land and therefore have fewer species present.
*	Crucially, some sites chosen for tree planting may have priority species or habitat indicators present that are not shown on the maps, as they have not been visited or well recorded by BSBI recorders. For these sites, field surveys will be required. Therefore, these maps should be used with caution and local site experts should be consulted wherever possible. 
*	The status of priority species is currently flagged as those which are rare and/or threatened at the GB scale. Use of equivalent species at the England-level would identify a different pool of priority species. However, a review of the species list demonstrated that the majority of species are included within the positive indicator lists, although with slightly different threatened and rare statuses noted. Globally or highly restricted endemic species were not reviewed, although many are likely to fall within the national threatened, rare and scarce species lists.
* Similarly, some local priority species may not necessarily mean they are a national priority species, and so local environmental teams may be aware of other key species to consider when planning actions on the ground.
*	The variation in survey coverage by BSBI recorders will also greatly impact the coverage of priority species and habitat indicators on these maps; wherever possible we have attempted to alleviate some of this bias through the localised benchmarking used to produce the botanical value map. Survey coverage also varies with the capture resolution of recording. Here we have only used records captured at the monad scale (1 x 1 km) however this does present a notable difference with areas which have historically captured data at lower resolution, with a defined difference notable between Cornwall and Devon.



## Conclusions 

Out of the trialled methods, the 25 km moving neighbourhood yielded the greatest number of monads categorised as being of high and moderate botanical value. Discussions with the Natural England habitat specialists and BSBI botanical experts agreed this methodology seemed most suitable for alleviating some of the spatial bias in indicator presence and was in line with previous studies adopting a similar methodology.
